\bf Tuning parameter selection for a penalized estimator of species richness

### 2019-08-06 rre_manuscript/paper.tex
round two of editing; add refs; JCGS style file

\documentclass[oupdraft]{bio}
\usepackage{url}

\usepackage{amsmath}
\usepackage{xparse}
\usepackage{mathtools}
\usepackage[table,dvipsnames]{xcolor}

@@

\newboolean{DEBUG}
\setboolean{DEBUG}{true}

\newrgbcolor{amycolor}{.7 .1 .1}
\ifthenelse {\boolean{DEBUG}}
{\newcommand{\amy}[1]{{\bf \amycolor{[\@Amy: #1]}}}}
{\newcommand{\amy}[1]{}}

\newrgbcolor{alexcolor}{.1 .7 .7}
\ifthenelse {\boolean{DEBUG}}
{\newcommand{\alex}[1]{{\bf \alexcolor{[\@Alex: #1]}}}}
{\newcommand{\alex}[1]{}}

\newrgbcolor{editcolor}{.1 .5 .7}
\ifthenelse {\boolean{DEBUG}}
{\newcommand{\edit}[1]{{\bf \editcolor{[\@edit: #1]}}}}
{\newcommand{\edit}[1]{}}

@@

% Title of paper
\title{Tuning parameter selection for a penalized estimator of species richness}

% List of authors, with corresponding author marked by asterisk
\author{ALEX PAYNTER, AMY D WILLIS$^\ast$\\[4pt]
% Author addresses
\textit{Department of Biostatistics,
University of Washington,
Health Sciences Building, Box 357232,
1705 NE Pacific St., Seattle, WA 98195}
% \\[2pt]
% % E-mail address for correspondence
% {payntera@uw.edu, adwillis@uw.edu}
}

% Running headers of paper:
\markboth%
% First field is the short list of authors
{A. Paynter and A.D. Willis}
% Second field is the short title of the paper
{Tuning parameter selection for species richness}

\maketitle

% Add a footnote for the corresponding author if one has been
% identified in the author list
\footnotetext{adwillis@uw.edu}

\begin{abstract}
{Our goal is to estimate the true number of classes in a population, called the species richness.  We consider the case where multiple frequency count tables have been collected from a homogeneous population, and investigate a penalized maximum likelihood estimator under a negative binomial model.
Because high probabilities of unobserved classes increase the variance of species richness estimates,
our method penalizes the probability of a class being unobserved, since high probabilities of unobserved classes increase the variance of species richness estimates. Tuning the penalization parameter is challenging because the true species richness is never known, and so we propose and validate four novel methods for tuning the penalization parameter.  We illustrate and contrast the performance of the proposed methods by estimating the strain-level microbial diversity of Lake Champlain over 3 consecutive years.}
\end{abstract}

@@

In ecology, species richness is a quantitative measurement of ecosystem diversity.  We focus on the specific application of estimating microbial diversity, that is, the number of strains of bacteria present in a population (e.g., a lake microenvironment or in an individual's oral cavity).
Microbial diversity is often linked with ecosystem health, such as in the vaginal microbiome (where high diversity is associated with infection \citep{Oakley:2008uo}) and in the gut microbiome (where high diversity is associated with healthy metabolism \citep{Morgan:2012bm,Minot:2019hc}).
Microbial abundance data typically contain many species observed infrequently (\textit{rare species}) and some which are observed a large number of times (\textit{abundant species}).  In data with both many rare and many abundant species, richness estimation methods often have high variance \citep{wang_2005,willis_2015}, motivating our regularised approach to estimation.

@@

This paper is organized as follows: We review species richness estimation in Section \ref{sec:literature_review}.  In Section \ref{sec:fixed_lambda} we describe our extension to biological replicates, and in Section \ref{sec:efficacy_sims} we establish the utility of penalization via simulations. In Section \ref{sec:tuning_proposals} we propose several novel methods for tuning the penalization parameter, and in Section \ref{sec:tuning_simulations}
we evaluate our proposals. \amy{Section \ref{sec:data_analysis} applies our methods to microbiome data on Lake Champlain...} Section \ref{sec:discussion} closes with an expanded discussion of our results and suggested directions for future work. Software implementing the methods is available in the \url{R} package \url{rre}, available at \url{github.com/statdivlab/rre}, and code to reproduce the simulation results is available at \url{github.com/statdivlab/rre_sims}.

@@

The classical model for estimating species abundance is a Poisson mixture model.  Under this model the number of times we observe an species $i$ is $X_i \sim \text{Poisson}(\Lambda)$ where $\Lambda$ is a random variable distributed according to some mixing distribution.
Many mixing distributions have been proposed, including by  \cite{bulmer_1974,ord_1986,norris_1998} (see \cite{bunge_1993} for a comprehensive review). While our proposal may be easily generalized to other mixing distributions, in this paper we consider the Gamma  distribution for $\Lambda$, first considered by \citet{fisher_1943}.

Let $f_k = \#\{i: X_i = k\}$ be the number of classes observed $k$ times, called the \textit{frequencies} or \textit{frequency counts}.  Then the set $\{f_k\}_{k \geq 1}$ is a \textit{frequency count table}, a common way to represent species abundance data. Because $C = f_0 + f_1 + f_2 + \dots = f_0 + c$, the species problem can be framed as predicting $f_0$ to obtain the species richness estimate $\hat{C} = \hat{f}_0 + c$.

Let $\Lambda \sim \text{Gamma}(\alpha, \delta)$ with distribution function $f(\lambda) = \delta^\alpha \Gamma (\alpha)^{-1} \lambda^{\alpha - 1} e^{-\delta \lambda}$, and write $\eta = (\alpha, \delta)$.
% Then the distribution of $X_i$ under a gamma-Poisson model can written as a negative-binomial distribution,
% \begin{align}
% p_{\eta}(x) = \frac{\Gamma (x + \alpha)}{\Gamma (\alpha) x!} \left(\frac{\delta}{1+\delta} \right)^\alpha \left( \frac{1}{1+ \delta} \right)^x.
% \end{align}
% Note that we are interested in estimating the parameters of a negative binomial
% a zero-truncated negative binomial
%
There exist both frequentist \citep{fisher_1943} and Bayesian \citep{efron_1976,barger_2010} approaches to parameter estimation under this model. We will focus on frequentist maximum likelihood in this paper.
\amy{introduce Gamma later?}

@@

In Section \ref{sec:tuning_proposals} we will require an objective function defined in terms of a subset of indices of the frequency count tables, so we define it that way now.  Let $J \subseteq \{1, \dots, r\}$.  $J$ indicates the data being used in the evaluation of the objective function, with $J = \{1, \dots r\}$ meaning we use all replicates.  Let $\ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right)$ be the unpenalized log likelihood for replicate $j$.
Using the fact that the draws are independent, our objective function for a fixed $\lambda \geq 0$ is the sum of the log likelihoods for each individual replicate.
We define the \textit{penalized log likelihood} for multiple samples as
\begin{align}
\mathcal{O}_{\lambda}(C, \eta; J) :=& \sum_{j \in J} \ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right) - \lambda\log p_{\eta}(0) \\
 =& \sum_{j \in J} \biggl[ \log C! - \log (C-c)! + (C-c) \log p_\eta(0)  \nonumber \\
 & \qquad  - \sum_{k \geq 1} \log{} f_{kj}! + \sum_{k \geq 1} f_{kj} \log p_\eta(k) \biggr] - \lambda \log p_\eta(0)    \label{eq:objective}
\end{align}
By fixing $\lambda = 0$ we obtain the \textit{unpenalized log likelihood} for all replicates $j \in J$.  We define $\widehat{C}_{\lambda}$ to be the penalized maximum likelihood estimate for based on tuning parameter $\lambda$, that is,
\begin{equation}
\left(\widehat{C}_{\lambda},  \widehat{\eta}_{\lambda} \right) = \argmax_{C, \, \eta}  \mathcal{O}_{\lambda} \left(C, \eta ; \{1, \dots , r\} \right) \label{eq:ccc_hat_lambda}
\end{equation}
where $\widehat{\eta}_{\lambda}$ is a nuisance parameter.

We use root-mean-square error as the primary criterion for evaluating the performance of estimators in simulations.  Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \eqref{eq:ccc_hat_lambda} obtained in simulation number $s$ with  $\lambda$.  Then we define
\begin{equation}
 \text{RMSE}\left( \widehat{C}_{\lambda} \right) = \sqrt{ \frac{1}{n_{\text{sim}}} \sum_{s=1}^{n_{\text{sim}}} \left( \widehat{C}_{\lambda \langle s \rangle} - C\right)^2 }
\end{equation}

To maximize the likelihood we will use the direct approach, rather than the profile and conditional approaches reviewed in Section \ref{sec:literature_review}.  While it may be faster to use the profile or conditional approach, our investigations revealed that the results are not always the same in finite samples.
Since our objective is to evaluate approaches to tuning, we place priority on the accuracy of optimization over speed.

@@

\section{Efficacy of Penalization}
\label{sec:efficacy_sims}
\subsection{Simulation setup}

Before evaluating potential tuning parameter selection methods, we first establish that penalization can improve richness estimation.  \citet{wang_2005} established that it can improve estimates for $r = 1$, this needs to be verified for $r> 1$.
We propose simulating data under the gamma-Poisson model and maximizing the objective function with a set of fixed $\lambda$ values.  If there exists some $\lambda > 0$ where RMSE($\widehat{C}_{\lambda}$) is less than RMSE($\widehat{C}_0)$, we conclude that penalization is effective for the given data generating mechanism.

We simulated using $C = 1000$, and $r \in \{6, 10, 14\}$.  We found that $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$  provided a good fit to the data of  \citet{walsh2014restricted}, and $\eta = (10^{-2}, 10^{-5})$ provided a good fit to the data of \citet{tromas_2017}.
For completeness we also investigated nearby $\eta$ values: $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.
% To select $\eta$, we investigated the unregularised
%  two microbial datasets
% so choosing appropriate simulation values of $\eta$ is crucial in targeting our approach to microbial data.   We used two microbial datasets as motivating examples.  \citet{walsh2014restricted} provide a dataset from soil in a Swiss apple orchard and we found $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$ produced a frequency count table with a similar structure to their data.  The second example from \citet{tromas_2017} is a sample of cyanobacteria from Lake Champlain, and we found $\eta = (10^{-2}, 10^{-5})$ mimicked the structure of their data well.  We note that is possible that the gamma-Poisson model is not the best fit for these examples, but our goal is only to choose $\eta$ in an appropriate range.  Two nearby $\eta$ values are included, $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.  These add variety to our simulations in this section and context for our simulations in Section \ref{sec:tuning_simulation_2}.  Under each combination of $\eta$ and $r$ 100 simulations were completed.
In Table \ref{tab:eta_intuition} we provide a few summary measures for frequency count tables generated under each choice of $\eta$.  This characterizes the $\eta$ values by the proportion of unobserved, rare or abundant species we would expect to see. We return to a discussion of the different data structures in Section \ref{sec:tuning_simulations}.
For each combination of $\eta$ and $r$, we performed 100 simulations.
We investigated the grid $\lambda \in \{0, 5, \dots , 120\}$, and found that this grid was sufficiently expansive to ensure that the RMSE-minimising $\lambda$ was not on the boundary of the grid
(see Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}).

@@

% latex table generated in R 3.5.1 by xtable 1.8-2 package
% Thu May 23 22:11:47 2019
\begin{table}[t]
\tblcaption{The expected proportion of unobserved ($k=0$), singleton ($k=1$), rare ($k=1,2,3$), and abundant  ($k \geq 10$) species for 4 choices of $\eta$.  We also give the expected frequency count of the most abundant species when $C = 1000$.
\label{tab:eta_intuition}}
\centering
\footnotesize
\begin{tabular}{rcccc}
  \hline
 $\eta = $ & ($10^{-1}$, $10^{-1}$) & ($10^{-2}$, $10^{-5}$) & ($10^{-1}$, $10^{-3}$) & ($10^{-1}$, $10^{-5}$) \\
  \hline
  Proportion unobserved ($p_{\eta}(0)$) & 0.787 & 0.891 & 0.501 & 0.316 \\
  Proportion singletons ($p_{\eta}(1)$) & 0.072 & 0.009 & 0.050 & 0.032 \\
  Proportion rare ($\sum_{k=1}^3 p_{\eta}(k)$) & 0.130 & 0.016 & 0.097 & 0.061 \\
  Proportion abundant ($\sum_{k=10}^{\infty} p_{\eta}(k)$) & 0.028 & 0.083 & 0.340 & 0.583 \\
Expected max abundance ($\mathbb{E}\left[k_{\text{max}}\right]$) & $4.1 \times 10^{1}$ & $2.0 \times 10^{5}$ & $4.0 \times 10^{3}$ & $3.9 \times 10^{5}$ \\
   \hline
\end{tabular}
\normalsize
\end{table}

\begin{table}[h]
\tblcaption{The penalized maximum likelihood estimate of $C$ has lower RMSE for all investigated choices of $\eta$ and $r$ under a Gamma-mixed Poisson model for species abundances based on 100 simulations for each choice of $\eta$ and $r$.
$\lambda_{\text{opt}}$ is the value of $\lambda$ which produced the lowest RMSE.
$\widehat{C}_0$ is the $C$ estimate with $\lambda = 0$, and $\widehat{C}_{\lambda_{\text{opt}}}$ is the $C$ estimate using $\lambda = \lambda_{\text{opt}}$.
\label{tab:fixed_lambda_results}}
\centering
\begin{tabular}{ccccc}
  \hline
$\eta = (\alpha, \delta)$ & $r$ & RMSE ($\widehat{C}_{0}$) & $\lambda_{opt}$ & RMSE ($\widehat{C}_{\lambda_{ \text{opt}}}$) \\
  \hline
$\left(10^{-1}, 10^{-1}\right)$ &   6 & 796.90 &  10 & 326.50 \\
$\left(10^{-1}, 10^{-1}\right)$ &  10 & 527.53 &  15 & 286.27 \\
  $\left(10^{-1}, 10^{-1}\right)$ &  14 & 337.31 &  10 & 235.51 \\
  $\left(10^{-2}, 10^{-5}\right)$ &   6 & 735.95 &  15 & 516.37 \\
  $\left(10^{-2}, 10^{-5}\right)$ &  10 & 700.21 &  20 & 456.04 \\
  $\left(10^{-2}, 10^{-5}\right)$ &  14 & 666.55 &  20 & 470.35 \\
  $\left(10^{-1}, 10^{-3}\right)$ &   6 & 200.09 &  20 & 156.29 \\
  $\left(10^{-1}, 10^{-3}\right)$ &  10 & 213.22 &  55 & 142.64 \\
  $\left(10^{-1}, 10^{-3}\right)$ &  14 & 243.42 &  55 & 148.26 \\
  $\left(10^{-1}, 10^{-5}\right)$ &   6 & 401.17 &  55 & 147.13 \\
  $\left(10^{-1}, 10^{-5}\right)$ &  10 & 283.91 &  25 & 137.04 \\
  $\left(10^{-1}, 10^{-5}\right)$ &  14 & 415.19 &  70 & 126.72 \\
   \hline
\end{tabular}
\end{table}

@@

The results of the simulation are shown in Table \ref{tab:fixed_lambda_results}.  For every parameter choice a reduction in RMSE($\widehat{C}_{\lambda}$) is found for some $\lambda > 0$.  Depending on simulation parameters the reduction varied from 22\% to 70\% compared to RMSE($\widehat{C}_{0}$) for the best $\lambda$.  These results show that for a variety of $(\eta, r)$ choices, penalization improves species richness estimation.
We also see that the optimal $\lambda$ value varies considerably over $(\eta, r)$, ranging from 10 to 70.  This suggests that there is not a universally appropriate $\lambda$ choice. Therefore, tuning $\lambda$ to the sample appears desirable. We note that the optimal choice of $\lambda$ is not consistent either for fixed $r$ and variable $\eta$, nor for fixed $\eta$ and variable $r$.

In order to demonstrate the effect on $\widehat{C}_{\lambda}$ as we increase $\lambda$, we plot the results for two $\eta$ values in Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}.  For small values of $\lambda$ we acquire lower variance in the estimates while incurring some negative bias, and the RMSE improves as a result.  Eventually the bias difference dominates and the RMSE increases as $\lambda \to \infty$.

@@

We have established that penalization can improve richness estimates, but different abundance structures ($\eta$) require different penalization parameters ($\lambda$) to minimize RMSE.  Therefore a method which can tune the parameter based on a sample is warranted.  We propose several novel methods in this section.  Each is evaluated via simulation in Section \ref{sec:tuning_simulations}.

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib, bbm}
\usepackage{url} % not crucial - just used below for the URL

\usepackage{amsthm}
\usepackage{lipsum}

\usepackage{xparse}
\usepackage{mathtools}
\usepackage[table,dvipsnames]{xcolor}

@@

\newtheoremstyle{break}
  {\topsep}{\topsep}%
  {\itshape}{}%
  {\bfseries}{}%
  {\newline}{}%
\theoremstyle{break}
\newtheorem{theorem}{Theorem}

\newtheoremstyle{break}
  {\topsep}{\topsep}%
  {\itshape}{}%
  {\bfseries}{}%
  {\newline}{}%
\theoremstyle{break}
\newtheorem{prop}{Proposition}

\newboolean{DEBUG}
\setboolean{DEBUG}{true}
\newrgbcolor{amycolor}{.7 .1 .1}
\ifthenelse {\boolean{DEBUG}}
{\newcommand{\amy}[1]{{\bf \amycolor{[\@Amy: #1]}}}}
{\newcommand{\amy}[1]{}}

\RequirePackage[OT1]{fontenc}
\RequirePackage{amsthm,amsmath, bbm, xspace, graphics, graphicx}
%\RequirePackage{breqn}
%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}

\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{-.3in}%
\addtolength{\topmargin}{-.8in}%

@@

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if1\blind
{
  \title{\bf Tuning parameter selection for a penalized estimator of species richness}
  \author{Alex Paynter and Amy D Willis\\ \\
  Department of Biostatistics, University of Washington,\\
  Health Sciences Building, Box 357232,\\
  1705 NE Pacific St., Seattle, WA 98195 \\ \\
  adwillis@uw.edu}

  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Tuning parameter selection for a penalized estimator of species richness}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
  Our goal is to estimate the true number of classes in a population, called the species richness.  We consider the case where multiple frequency count tables have been collected from a homogeneous population, and investigate a penalized maximum likelihood estimator under a negative binomial model.
  Because high probabilities of unobserved classes increase the variance of species richness estimates,
  our method penalizes the probability of a class being unobserved. Tuning the penalization parameter is challenging because the true species richness is never known, and so we propose and validate four novel methods for tuning the penalization parameter.  We illustrate and contrast the performance of the proposed methods by estimating the strain-level microbial diversity of Lake Champlain over 3 consecutive years.
\end{abstract}

\noindent%
{\it Keywords: diversity, biodiversity, microbiome, regularization, maximum likelihood}
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

% % List of authors, with corresponding author marked by asterisk
% \author{ALEX PAYNTER, AMY D WILLIS$^\ast$\\[4pt]
% % Author addresses
% \textit{Department of Biostatistics,
% University of Washington,
% Health Sciences Building, Box 357232,
% 1705 NE Pacific St., Seattle, WA 98195}
% % \\[2pt]
% % % E-mail address for correspondence
% % {payntera@uw.edu, adwillis@uw.edu}
% }
%
% % Running headers of paper:
% \markboth%
% % First field is the short list of authors
% {A. Paynter and A.D. Willis}
% % Second field is the short title of the paper
% {Tuning parameter selection for species richness}
%
% \maketitle
%
% % Add a footnote for the corresponding author if one has been
% % identified in the author list
% \footnotetext{adwillis@uw.edu}

@@

In ecology, species richness is a quantitative measurement of ecosystem diversity.  We focus on the specific application of estimating microbial diversity, that is, the number of strains of bacteria present in a population (e.g., a lake microenvironment or in an individual's oral cavity).
Microbial diversity is often linked with ecosystem health, such as in the vaginal microbiome (where high diversity is associated with infection \citep{Oakley:2008uo}) and in the gut microbiome (where high diversity is associated with healthy metabolism \citep{Morgan:2012bm,Minot:2019hc}).
Microbial abundance data typically contain many species observed infrequently (\textit{rare species}) and some which are observed a large number of times (\textit{abundant species}).  In data with both many rare and many abundant species, richness estimation methods often have high variance \citep{wang_2005,willis_2015}, motivating our regularized approach to estimation.

@@

This paper is organized as follows: We review species richness estimation in Section \ref{sec:literature_review}.  In Section \ref{sec:fixed_lambda} we describe our extension to biological replicates, and in Section \ref{sec:efficacy_sims} we establish the utility of penalization via simulations. In Section \ref{sec:tuning_proposals} we propose several novel methods for tuning the penalization parameter, and in Section \ref{sec:tuning_simulations}
we evaluate our proposals. In Section \ref{sec:data_analysis} we apply our methods to estimate microbial diversity in Lake Champlain, and Section \ref{sec:discussion} closes with a discussion of our results and suggested directions for future work. Software implementing the methods is available in the \url{R} package \url{rre} (\textit{regularized richness estimation}), available at \url{github.com/statdivlab/rre}, and code to reproduce the simulation results is available at \url{github.com/statdivlab/rre_sims}.

@@

The classical model for estimating species abundance is a Poisson mixture model.  Under this model the number of times we observe species $i$ is $X_i \sim \text{Poisson}(\Lambda)$ where $\Lambda$ is a random variable distributed according to some mixing distribution.
Many mixing distributions have been proposed, including by  \cite{bulmer_1974,ord_1986,norris_1998} (see \cite{bunge_1993} for a comprehensive review). While our proposal may be easily generalized to other mixing distributions, in this paper we consider the Gamma  distribution for $\Lambda$, first considered by \citet{greenwood_yule} and \citet{fisher_1943}.
Note that we only observe $X_i | X_i > 0$, and therefore we are interested in estimating the parameters of a Poisson mixture model using \textit{zero-truncated} Poisson-mixed data.

Let $f_k = \#\{i: X_i = k\}$ be the number of classes observed $k$ times, called the \textit{frequency counts}.  Then the set $\{f_k\}_{k \geq 1}$ is a \textit{frequency count table}, a common way to represent species abundance data. Because $C = f_0 + f_1 + f_2 + \dots = f_0 + c$, the species problem can be framed as predicting $f_0$ given $f_1, f_2, \ldots$ to obtain a species richness estimate $\hat{C} = \hat{f}_0 + c$.

Let $\Lambda \sim \text{Gamma}(\alpha, \delta)$ with distribution function $f(\lambda) = \delta^\alpha \Gamma (\alpha)^{-1} \lambda^{\alpha - 1} e^{-\delta \lambda}$, and write $\eta = (\alpha, \delta)$.
% Then the distribution of $X_i$ under a gamma-Poisson model can written as a negative-binomial distribution,
% \begin{align}
% p_{\eta}(x) = \frac{\Gamma (x + \alpha)}{\Gamma (\alpha) x!} \left(\frac{\delta}{1+\delta} \right)^\alpha \left( \frac{1}{1+ \delta} \right)^x.
% \end{align}
% Note that we are interested in estimating the parameters of a negative binomial
% a zero-truncated negative binomial
%
There exist both frequentist \citep{fisher_1943,chao_2002} and Bayesian \citep{efron_1976,barger_2010} approaches to parameter estimation under this model. We will focus on frequentist maximum likelihood in this paper.

@@

In Section \ref{sec:tuning_proposals} we will require an objective function defined in terms of a subset of indices of the frequency count tables, so we define it that way now.  Let $J \subseteq \{1, \dots, r\}$.  $J$ indicates the data being used in the evaluation of the objective function, with $J = \{1, \dots r\}$ meaning we use all replicates.  Let $\ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right)$ be the unpenalized log likelihood for replicate $j$.
Using the fact that the draws are independent, our objective function for a fixed $\lambda \geq 0$ is the sum of the log likelihoods for each individual replicate.
We define the \textit{penalized log likelihood} for multiple samples as
\begin{align}
\mathcal{O}_{\lambda}(C, \eta; J) :=& \sum_{j \in J} \ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right) - \lambda\log p_{\eta}(0) \\
 =& \sum_{j \in J} \biggl[ \log C! - \log (C-c)! + (C-c) \log p_\eta(0)  \nonumber \\
 & \qquad  - \sum_{k \geq 1} \log{} f_{kj}! + \sum_{k \geq 1} f_{kj} \log p_\eta(k) \biggr] - \lambda \log p_\eta(0)    \label{eq:objective}
\end{align}
By fixing $\lambda = 0$ we obtain the unpenalized log likelihood for the set $J$.  We define $\widehat{C}_{\lambda}$ to be the penalized maximum likelihood estimate of $C$ based on tuning parameter $\lambda$, that is,
\begin{equation}
\left(\widehat{C}_{\lambda},  \widehat{\eta}_{\lambda} \right) = \argmax_{C, \, \eta}  \mathcal{O}_{\lambda} \left(C, \eta ; \{1, \dots , r\} \right) \label{eq:ccc_hat_lambda}
\end{equation}
where $\widehat{\eta}_{\lambda}$ is a nuisance parameter.

To maximize the likelihood we will use the direct approach, rather than the profile and conditional approaches reviewed in Section \ref{sec:literature_review}.  While it may be faster to use the profile or conditional approach, the results are not always the same in finite samples.
Since our objective is to evaluate approaches to tuning, we place priority on the accuracy of optimization over speed of optimization.

@@

\section{Evaluating penalized species richness estimates}
\label{sec:efficacy_sims}
\subsection{Simulation setup}

Before evaluating potential tuning parameter selection methods, we first establish that penalization can improve richness estimation.  \citet{wang_2005} established that it can improve estimates for $r = 1$, this needs to be verified for $r> 1$.

\amy{note that it's not a penalty, it's a bonus}

We use square root mean square error (RMSE) as the primary criterion for evaluating the performance of estimators.  Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \eqref{eq:ccc_hat_lambda} obtained in simulation number $s$ with  $\lambda$.  Then we define
\begin{equation}
 \text{RMSE}\left( \widehat{C}_{\lambda} \right) = \sqrt{ \frac{1}{n_{\text{sim}}} \sum_{s=1}^{n_{\text{sim}}} \left( \widehat{C}_{\lambda \langle s \rangle} - C\right)^2 }.
\end{equation}
In this simulation, if there exists some $\lambda > 0$ for which RMSE($\widehat{C}_{\lambda}$) is less than RMSE($\widehat{C}_0)$, we conclude that penalization is effective.

We simulated using $C = 1000$, and $r \in \{6, 10, 14\}$.  We found that $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$  provided a good fit to the data of  \citet{walsh2014restricted}, and $\eta = (10^{-2}, 10^{-5})$ provided a good fit to the data of \citet{tromas_2017} (see Section \ref{sec:data_analysis}).
For completeness we also investigated nearby $\eta$ values: $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.
% To select $\eta$, we investigated the unregularised
%  two microbial datasets
% so choosing appropriate simulation values of $\eta$ is crucial in targeting our approach to microbial data.   We used two microbial datasets as motivating examples.  \citet{walsh2014restricted} provide a dataset from soil in a Swiss apple orchard and we found $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$ produced a frequency count table with a similar structure to their data.  The second example from \citet{tromas_2017} is a sample of cyanobacteria from Lake Champlain, and we found $\eta = (10^{-2}, 10^{-5})$ mimicked the structure of their data well.  We note that is possible that the gamma-Poisson model is not the best fit for these examples, but our goal is only to choose $\eta$ in an appropriate range.  Two nearby $\eta$ values are included, $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.  These add variety to our simulations in this section and context for our simulations in Section \ref{sec:tuning_simulation_2}.  Under each combination of $\eta$ and $r$ 100 simulations were completed.
In Table \ref{tab:eta_intuition} we provide  summary measures for frequency count tables generated under each choice of $\eta$.  This characterizes the $\eta$ values by the proportion of unobserved, rare or abundant species we would expect to see. We return to a discussion of the different data structures in Section \ref{sec:tuning_simulations}.
For each combination of $\eta$ and $r$, we performed 100 simulations.
We investigated the grid $\lambda \in \{0, 5, \dots , 120\}$, and found that this grid was sufficiently expansive to ensure that the RMSE-minimising $\lambda$ was not on the boundary of the grid
(see Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}).

@@

% latex table generated in R 3.5.1 by xtable 1.8-2 package
% Thu May 23 22:11:47 2019
\begin{table}[t]
\caption{The expected proportion of unobserved ($k=0$), singleton ($k=1$), rare ($k=1,2,3$), and abundant  ($k \geq 10$) species for 4 choices of $\eta$.  We also give the expected frequency count of the most abundant species when $C = 1000$.
\label{tab:eta_intuition}}
\centering
\footnotesize
\begin{tabular}{rcccc}
  \hline
 $\eta = $ & ($10^{-1}$, $10^{-1}$) & ($10^{-2}$, $10^{-5}$) & ($10^{-1}$, $10^{-3}$) & ($10^{-1}$, $10^{-5}$) \\
  \hline
  Proportion unobserved ($p_{\eta}(0)$) & 0.787 & 0.891 & 0.501 & 0.316 \\
  Proportion singletons ($p_{\eta}(1)$) & 0.072 & 0.009 & 0.050 & 0.032 \\
  Proportion rare ($\sum_{k=1}^3 p_{\eta}(k)$) & 0.130 & 0.016 & 0.097 & 0.061 \\
  Proportion abundant ($\sum_{k=10}^{\infty} p_{\eta}(k)$) & 0.028 & 0.083 & 0.340 & 0.583 \\
Expected max abundance ($\mathbbm{E}\left[k_{\text{max}}\right]$) & $4.1 \times 10^{1}$ & $2.0 \times 10^{5}$ & $4.0 \times 10^{3}$ & $3.9 \times 10^{5}$ \\
   \hline
\end{tabular}
\normalsize
\end{table}

\begin{table}[h]
\caption{The penalized maximum likelihood estimate of $C$ has lower RMSE for all investigated choices of $\eta$ and $r$ under a Gamma-mixed Poisson model for species abundances based on 100 simulations for each choice of $\eta$ and $r$.
$\lambda_{\text{opt}}$ is the value of $\lambda$ which produced the lowest RMSE.
$\widehat{C}_0$ is the $C$ estimate with $\lambda = 0$, and $\widehat{C}_{\lambda_{\text{opt}}}$ is the $C$ estimate using $\lambda = \lambda_{\text{opt}}$.
\label{tab:fixed_lambda_results}}
\centering
\begin{tabular}{ccccc}
  \hline
$\eta = (\alpha, \delta)$ & $r$ & RMSE ($\widehat{C}_{0}$) & $\lambda_{opt}$ & RMSE ($\widehat{C}_{\lambda_{ \text{opt}}}$) \\
  \hline
$\left(10^{-1}, 10^{-1}\right)$ &   6 & 796.90 &  10 & 326.50 \\
$\left(10^{-1}, 10^{-1}\right)$ &  10 & 527.53 &  15 & 286.27 \\
  $\left(10^{-1}, 10^{-1}\right)$ &  14 & 337.31 &  10 & 235.51 \\
  $\left(10^{-2}, 10^{-5}\right)$ &   6 & 735.95 &  15 & 516.37 \\
  $\left(10^{-2}, 10^{-5}\right)$ &  10 & 700.21 &  20 & 456.04 \\
  $\left(10^{-2}, 10^{-5}\right)$ &  14 & 666.55 &  20 & 470.35 \\
  $\left(10^{-1}, 10^{-3}\right)$ &   6 & 200.09 &  20 & 156.29 \\
  $\left(10^{-1}, 10^{-3}\right)$ &  10 & 213.22 &  55 & 142.64 \\
  $\left(10^{-1}, 10^{-3}\right)$ &  14 & 243.42 &  55 & 148.26 \\
  $\left(10^{-1}, 10^{-5}\right)$ &   6 & 401.17 &  55 & 147.13 \\
  $\left(10^{-1}, 10^{-5}\right)$ &  10 & 283.91 &  25 & 137.04 \\
  $\left(10^{-1}, 10^{-5}\right)$ &  14 & 415.19 &  70 & 126.72 \\
   \hline
\end{tabular}
\end{table}

@@

The results of the simulation are shown in Table \ref{tab:fixed_lambda_results}.  For every parameter choice a reduction in RMSE($\widehat{C}_{\lambda}$) was found for some $\lambda > 0$.  Depending on simulation parameters the reduction varied from 22\% to 70\% compared to RMSE($\widehat{C}_{0}$) for the best $\lambda$.  These results show that for a variety of $(\eta, r)$ choices, penalization improves species richness estimation.
We also see that the optimal $\lambda$ value varies considerably over $(\eta, r)$, ranging from 10 to 70.  This suggests that there is not a universally appropriate $\lambda$ choice. Therefore, tuning $\lambda$ to the sample appears desirable. We note that the optimal choice of $\lambda$ is not consistent either for fixed $r$ and variable $\eta$, nor for fixed $\eta$ and variable $r$.

In order to demonstrate the effect on $\widehat{C}_{\lambda}$ as we increase $\lambda$, we plot the results for two $\eta$ values in Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}.  For small values of $\lambda$ we acquire lower variance in the estimates while incurring some negative bias, and the RMSE improves as a result.  Eventually the bias term dominates and the RMSE increases as $\lambda \to \infty$.

@@

We have established that penalization can improve richness estimates, but different abundance structures ($\eta$) require different values of $\lambda$ to minimize RMSE. We therefore develop methods to tune $\lambda$ based on a sample.  We propose several novel methods in this section.  Each is evaluated in Section \ref{sec:tuning_simulations}.
